Assessing the potential of Jacobian-free Newton-Krylov methods for cell-centred finite volume solid mechanics

### 2025-01-17 article.tex
Add distorted hex mesh to MMS

The manufactured displacement solution is applied at the domain's boundaries (prescribed displacements), and inertial effects are neglected.
Three mesh types are examined:
%(i) \emph{regular} hexahedra, (ii) \emph{distorted} hexahedra, (iii) tetrahedra, and (iv) polyhedra (shown in Figure \ref{fig:mms_mesh}).
(i) hexahedra, (ii) tetrahedra, and (iii) polyhedra (shown in Figure \ref{fig:mms_mesh}).
The hexahedral meshes are created using the OpenFOAM \texttt{blockMesh} utility, while the tetrahedral meshes are created using Gmsh \citep{geuzaine2009gmsh}.
To create the polyhedral meshes, the tetrahedra meshes are converted to their dual polyhedral representations using the OpenFOAM \texttt{polyDualMesh} utility.
%The distorted hexahedral meshes are created by perturbing the points of the regular hexahedra by a random vector of magnitude less than 0.3 times the local minimum edge length.
%Points on the corners and edges of the cube domain are not moved, while the remaining boundary points are limited to slide along the boundary.
%During perturbation, if needed, the local point perturbation is iteratively reduced to ensure the local distorted cell volumes are not less than 10\% of their original volume and the maximum face non-orthogonality is less than $70^\circ$.
%Consequently, the meshes can be considered moderately distorted.
%Appendix \ref{app:meshes} includes images of all three cell types for both regular and distorted configurations.
Starting from an initial mesh spacing of $0.04$ m, six meshes of each type are created by successively halving the spacing.
The cell numbers for the hexahedral and polyhedral meshes are 125, $1\,000$, $8\,000$, $64\,000$, $512\,000$, and $4\,096\,000$, while the tetrahedral mesh cell counts are 384, $4\,374$, $41\,154$, $355\,914$, $2\,958\,234$, and $24\,118\,074$.
For the same average cell width, the cell counts show that the tetrahedral meshes have more cells by a factor of 3 to 6.
Table \ref{tab:mmsMesh} lists the average and maximum face non-orthogonality for the meshes of different cell types, where face non-orthogonality is defined as the angle made by the vector between the two adjacent cell centres across the common face and the face normal.
\begin{table}[htb]
    \centering
    \begin{tabular}{lll}
        \hline
        Cell type & Average & Maximum \\
        \hline
        Hexahedra & 0 &0  \\
        Tetrahedra & 32.71 & 35.26 \\
        Polyhedra & 30.65 & 38.45 \\
        \hline
    \end{tabular}
%    \begin{tabular}{l|cc|cc}
%        \hline
%        Cell type & \multicolumn{2}{c|}{Regular} & \multicolumn{2}{c}{Distorted} \\
%        & Average & Maximum & Average & Maximum \\
%        \hline
%        Hexahedra & 0 &0 & 13.10 & 52.51 \\
%        Tetrahedra & 32.73 & 35.26 & 34.83 & 69.96 \\
%        Polyhedra & 30.90 & 38.38 & 3 & 4 \\
%        \hline
%    \end{tabular}
    \caption{Average and maximum face non-orthogonality (in $^\circ$) for the meshes of different cell types and sizes}
    \label{tab:mmsMesh}
\end{table}

Figure \ref{fig:mms_disp_accuracy}(a) shows the displacement magnitude discretisation errors ($L_2$ and $L_\infty$) as a function of the average cell width for the three mesh types (hexahedral, tetrahedral and polyhedral), while Figure \ref{fig:mms_disp_accuracy}(b) shows the corresponding order of accuracy plots.
For ease of interpretation, the symbol shapes in the figures have been chosen to correspond to the cell shapes: a square for hexahedra, a triangle for tetrahedra and a pentagon for polyhedra.
The maximum ($L_\infty$) and average ($L_2$) displacement discretisation errors are seen to reduce at an approximately second-order rate for all mesh types, except for the $L_2$ error on the hexahedral meshes, which is approximately 2.3 on the finest grid.
%where, interestingly, the errors are, on average, smaller on the polyhedral meshes.
%The reason for this is unclear but may be due to the stencil of the structured polyhedra being more isotropic and containing close neighbours.
\begin{figure}[htbp]
	\centering
	\subfigure[Displacement magnitude discretisation errors]
	{
   		\includegraphics[width=0.45\textwidth]{figures/mms_dispErrors} 
   	}
	\subfigure[Displacement discretisation error order of accuracy]
	{
   		\includegraphics[width=0.45\textwidth]{figures/mms_disp_orderOfAccuracy}  
   	}
	\caption{Manufactured solution cube case: the accuracy and order of accuracy for displacement magnitude on the \emph{regular} meshes}
	\label{fig:mms_disp_accuracy}
\end{figure}

The predicted $\sigma_{xx}$ stress distribution for a hexahedral mesh with $512\,000$ cells is shown in Figure \ref{fig:mms_stress}(a).
The corresponding cell-wise $\sigma_{xx}$ error distribution is shown in Figure \ref{fig:mms_stress}(b).
The errors of greatest magnitude ($43$ kPa) occur at the boundaries, corresponding to where the local truncation error is higher. % due to the use of one-sided differencing.
\begin{figure}[htbp]
	\centering
	\subfigure[Predicted $\sigma_{xx}$ stress distribution]
	{
		\label{fig:mms_sxx}
   		\includegraphics[height=0.35\textwidth]{figures/mms_sxx} 
   	}
	\subfigure[Cell-wise $\sigma_{xx}$ error distribution]
	{
		\label{fig:mms_sxx_diff}
   		\includegraphics[height=0.35\textwidth]{figures/mms_sxx_diff}  
   	}
	\caption{Manufactured solution cube case: the predicted $\sigma_{xx}$ stress distribution on a cut-plane for the regular hexahedral mesh with $512\,000$ cells (left).}
	\label{fig:mms_stress}
\end{figure}
The discretisation errors in the stress magnitude as a function of cell size are shown in Figure \ref{fig:mms_stress_accuracy}(a), and the order of accuracy in Figure \ref{fig:mms_stress_accuracy}(b).
The order of accuracy for the average ($L_2$) stress error is seen to be approximately 1.5 for the hexahedral and polyhedral meshes.
In contrast, the average stress error order for the tetrahedral meshes is 1.
Similarly, the maximum ($L_\infty$) stress order of accuracy is seen to approach 1 for all three mesh types.
\begin{figure}[htbp]
	\centering
	\subfigure[Stress magnitude discretisation errors]
	{
   		\includegraphics[width=0.45\textwidth]{figures/mms_stressErrors} 
   	}
	\subfigure[Stress discretisation error order of accuracy]
	{
   		\includegraphics[width=0.45\textwidth]{figures/mms_stress_orderOfAccuracy}  
   	}
	\caption{Manufactured solution cube case: the accuracy and order of accuracy for stress magnitude}
	\label{fig:mms_stress_accuracy}
\end{figure}

@@

The true stress can be calculated from the second Piola–Kirchhoff stress as
\begin{eqnarray}
	\bb{\sigma} &=& \frac{1}{J} \bb{F} \cdot \bb{S} \cdot \bb{F}^{\text{T}}
\end{eqnarray}

\subsection{Neo-Hookean Hyperelasticity}
The definition of true (Cauchy) stress $\bb{\sigma}$ for neo-Hookean hyperelasticity can be given as
\begin{eqnarray} \label{eq:neoHook}
	\bb{\sigma}
	&=& \frac{\mu}{J} \, \text{dev} \left( \bar{\bb{b}} \right) + \frac{\kappa}{2} \frac{J^2 - 1}{J} \textbf{I}
\end{eqnarray}
where, once again, $\mu$ is the shear modulus, and $\kappa$ is the bulk modulus.
The bulk modulus can be expressed in term of the Young's modulus ($E$) and Poisson's ratio $\nu$ as
\begin{eqnarray}
	\kappa = \frac{E}{3(1 - 2\nu)}
\end{eqnarray}
The volume-preserving component of the elastic left Cauchy--Green deformation tensor is $\boldsymbol{b}$ is given as
\begin{eqnarray}
	\bar{\bb{b}} = J^{-2/3} \bb{b} = J^{-2/3} \bb{F} \cdot \bb{F}^{\text{T}}
\end{eqnarray}
In the limit of small deformations $\lVert \nabla \mathbf{u} \rVert  \ll 1$, neo-Hookean hyperelasticity (Equation \ref{eq:neoHook}) reduces to linear elasticity (Equation \ref{eq:linearElastic}).

\subsection{Guccione Hyperelasticity}
The \citet{Guccione1995} hyperelastic law defines the second Piola-Kirchhoff stress as
\begin{eqnarray}
	\boldsymbol{S}
		%&=&  \frac{\partial \Psi}{\partial \boldsymbol{E} } \notag \\
		&=& \frac{\partial Q}{\partial \boldsymbol{E} } \left( \frac{C}{2} \right) e^Q + \frac{\kappa}{2} \frac{J^2 - 1}{J} \textbf{I} 
\end{eqnarray}
where
\begin{eqnarray}
%	\Psi(I_1, I_2, I_4, I_5) &=& \frac{C}{2} \left( e^{Q(I_1, I_2, I_4, I_5)} - 1 \right) \\
	Q(I_1, I_2, I_4, I_5) &=& c_t I_1^2 - 2 c_t I_2 + (c_f - 2c_{fs} + c_t) I_4^2   + 2(c_{fs} - c_t) I_5 \\
	\frac{\partial Q}{\partial \boldsymbol{E} }
%		&=& \frac{\partial Q}{\partial I_1} \frac{\partial I_1}{\partial \boldsymbol{E} } + \frac{\partial Q}{\partial I_2} \frac{\partial I_2}{\partial \boldsymbol{E} }
%		+ \frac{\partial Q}{\partial I_4} \frac{\partial I_4}{\partial \boldsymbol{E} } + \frac{\partial Q}{\partial I_5} \frac{\partial I_5}{\partial \boldsymbol{E} } \notag \\
		&=&  2 c_t \boldsymbol{E} + 2(c_f - 2 c_{fs} + c_t)I_4 ( \boldsymbol{f_0} \otimes \boldsymbol{f_0} ) \notag \\
		&&+ 2(c_{fs} - c_t)\left[\boldsymbol{E} \cdot  (\boldsymbol{f_0} \otimes \boldsymbol{f_0}) + ( \boldsymbol{f_0} \otimes \boldsymbol{f_0} ) \cdot \boldsymbol{E} \right]
\end{eqnarray}
The scalars $C$, $c_f$, $c_{fs}$, and $c_t$ are material parameters, and invariants of the Green strain, $\boldsymbol{E} = \boldsymbol{F}^{\text{T}} \cdot \boldsymbol{F}$, are defined as
\begin{eqnarray}
	I_1 = \text{tr}(\boldsymbol{E}), \quad
	I_2 =  \frac{1}{2} \left[ \text{tr}^2(\boldsymbol{E}) - \text{tr}(\boldsymbol{E} \cdot \boldsymbol{E}) \right], \notag \\
	I_4 = \boldsymbol{E}  : \left( \boldsymbol{f_0} \otimes \boldsymbol{f_0} \right), \quad 
	I_5 = \left( \boldsymbol{E} \cdot \boldsymbol{E} \right): \left( \boldsymbol{f_0} \otimes \boldsymbol{f_0} \right)
\end{eqnarray}
with $\boldsymbol{f_0}$ representing the unit fibre directions in the initial configuration.

@@

>>>>

The manufactured displacement solution is applied at the domain's boundaries (prescribed displacements), and inertial effects are neglected.
Three mesh types are examined:
%(i) hexahedra, (ii) tetrahedra, and (iii) polyhedra (shown in Figure \ref{fig:mms_mesh}).
(i) \emph{regular} tetrahedra, (ii) \emph{regular} polyhedra (shown in Figure \ref{fig:mms_mesh}), (iii) \emph{regular} hexahedra, and (iv) \emph{distorted} hexahedra.
The second coarsest meshes are shown in Appendix \ref{app:meshes}.
The tetrahedral meshes are created using Gmsh \citep{geuzaine2009gmsh}, while the polyhedral meshes are created by converting the tetrahedral meshes to their dual polyhedral representations using the OpenFOAM \texttt{polyDualMesh} utility.
The regular hexahedral meshes are created using the OpenFOAM \texttt{blockMesh} utility, and the distorted hexahedral meshes are created by perturbing the points of the regular hexahedra by a random vector of magnitude less than 0.3 times the local minimum edge length.
%Points on the corners and edges of the cube domain are not moved, while the remaining boundary points are limited to slide along the boundary.
%During perturbation, if needed, the local point perturbation is iteratively reduced to ensure the local distorted cell volumes are not less than 10\% of their original volume and the maximum face non-orthogonality is less than $70^\circ$.
%Consequently, the meshes can be considered moderately distorted.
%Appendix \ref{app:meshes} includes images of all three cell types for both regular and distorted configurations.
Starting from an initial mesh spacing of $0.04$ m, six meshes of each type are created by successively halving the spacing.
The cell numbers for the hexahedral and polyhedral meshes are 125, $1\,000$, $8\,000$, $64\,000$, $512\,000$, and $4\,096\,000$, while the tetrahedral mesh cell counts are 384, $4\,374$, $41\,154$, $355\,914$, $2\,958\,234$, and $24\,118\,074$.
For the same average cell width, the cell counts show that the tetrahedral meshes have more cells by a factor of 3 to 6.
Table \ref{tab:mmsMesh} lists the average and maximum face non-orthogonality for the meshes of different cell types, where face non-orthogonality is defined as the angle made by the vector between the two adjacent cell centres across the common face and the face normal.
\begin{table}[htb]
    \centering
    \begin{tabular}{lll}
        \hline
        Cell type & Average & Maximum \\
        \hline
        Hexahedra & 0 &0  \\
        Tetrahedra & 32.71 & 35.26 \\
        Polyhedra & 30.65 & 38.45 \\
        \hline
    \end{tabular}
%    \begin{tabular}{l|cc|cc}
%        \hline
%        Cell type & \multicolumn{2}{c|}{Regular} & \multicolumn{2}{c}{Distorted} \\
%        & Average & Maximum & Average & Maximum \\
%        \hline
%        Hexahedra & 0 &0 & 13.10 & 52.51 \\
%        Tetrahedra & 32.73 & 35.26 & 34.83 & 69.96 \\
%        Polyhedra & 30.90 & 38.38 & 3 & 4 \\
%        \hline
%    \end{tabular}
    \caption{Average and maximum face non-orthogonality (in $^\circ$) for the meshes of different cell types and sizes}
    \label{tab:mmsMesh}
\end{table}

Figure \ref{fig:mms_disp_accuracy}(a) shows the displacement magnitude discretisation errors ($L_2$ and $L_\infty$) as a function of the average cell width for the three mesh types (hexahedral, tetrahedral and polyhedral), while Figure \ref{fig:mms_disp_accuracy}(b) shows the corresponding order of accuracy plots.
For ease of interpretation, the symbol shapes in the figures have been chosen to correspond to the cell shapes: a square for hexahedra, a triangle for tetrahedra and a pentagon for polyhedra.
The maximum ($L_\infty$) and average ($L_2$) displacement discretisation errors are seen to reduce at an approximately second-order rate for all mesh types, except for the $L_2$ error on the hexahedral meshes, which is approximately 2.3 on the finest grid.
%where, interestingly, the errors are, on average, smaller on the polyhedral meshes.
%The reason for this is unclear but may be due to the stencil of the structured polyhedra being more isotropic and containing close neighbours.
\begin{figure}[htbp]
	\centering
	\subfigure[Displacement magnitude discretisation errors]
	{
   		\includegraphics[width=0.45\textwidth]{figures/mms_dispErrors_v2} 
   	}
	\subfigure[Displacement discretisation error order of accuracy]
	{
   		\includegraphics[width=0.45\textwidth]{figures/mms_disp_orderOfAccuracy_v2}  
   	}
	\caption{Manufactured solution cube case: the accuracy and order of accuracy for displacement magnitude on the \emph{regular} meshes}
	\label{fig:mms_disp_accuracy}
\end{figure}

The predicted $\sigma_{xx}$ stress distribution for a hexahedral mesh with $512\,000$ cells is shown in Figure \ref{fig:mms_stress}(a).
The corresponding cell-wise $\sigma_{xx}$ error distribution is shown in Figure \ref{fig:mms_stress}(b).
The errors of greatest magnitude ($43$ kPa) occur at the boundaries, corresponding to where the local truncation error is higher. % due to the use of one-sided differencing.
\begin{figure}[htbp]
	\centering
	\subfigure[Predicted $\sigma_{xx}$ stress distribution]
	{
		\label{fig:mms_sxx}
   		\includegraphics[height=0.35\textwidth]{figures/mms_sxx} 
   	}
	\subfigure[Cell-wise $\sigma_{xx}$ error distribution]
	{
		\label{fig:mms_sxx_diff}
   		\includegraphics[height=0.35\textwidth]{figures/mms_sxx_diff}  
   	}
	\caption{Manufactured solution cube case: the predicted $\sigma_{xx}$ stress distribution on a cut-plane for the regular hexahedral mesh with $512\,000$ cells (left).}
	\label{fig:mms_stress}
\end{figure}
The discretisation errors in the stress magnitude as a function of cell size are shown in Figure \ref{fig:mms_stress_accuracy}(a), and the order of accuracy in Figure \ref{fig:mms_stress_accuracy}(b).
The order of accuracy for the average ($L_2$) stress error is seen to be approximately 1.5 for the hexahedral and polyhedral meshes.
In contrast, the average stress error order for the tetrahedral meshes is 1.
Similarly, the maximum ($L_\infty$) stress order of accuracy is seen to approach 1 for all three mesh types.
\begin{figure}[htbp]
	\centering
	\subfigure[Stress magnitude discretisation errors]
	{
   		\includegraphics[width=0.45\textwidth]{figures/mms_stressErrors_v2} 
   	}
	\subfigure[Stress discretisation error order of accuracy]
	{
   		\includegraphics[width=0.45\textwidth]{figures/mms_stress_orderOfAccuracy_v2}  
   	}
	\caption{Manufactured solution cube case: the accuracy and order of accuracy for stress magnitude}
	\label{fig:mms_stress_accuracy}
\end{figure}

@@

The true stress can be calculated from the second Piola–Kirchhoff stress as
\begin{eqnarray} \label{eq:S2sigma}
	\bb{\sigma} &=& \frac{1}{J} \bb{F} \cdot \bb{S} \cdot \bb{F}^{\text{T}}
\end{eqnarray}

\subsection{Neo-Hookean Hyperelasticity}
The definition of true (Cauchy) stress $\bb{\sigma}$ for neo-Hookean hyperelasticity can be given as
\begin{eqnarray} \label{eq:neoHook}
	\bb{\sigma}
	&=& \frac{\mu}{J} \, \text{dev} \left( \bar{\bb{b}} \right) + \frac{\kappa}{2} \frac{J^2 - 1}{J} \textbf{I}
\end{eqnarray}
where, once again, $\mu$ is the shear modulus, and $\kappa$ is the bulk modulus.
The bulk modulus can be expressed in terms of the Young's modulus ($E$) and Poisson's ratio $\nu$ as
\begin{eqnarray}
	\kappa = \frac{E}{3(1 - 2\nu)}
\end{eqnarray}
The volume-preserving component of the elastic left Cauchy--Green deformation tensor is $\boldsymbol{b}$ is given as
\begin{eqnarray}
	\bar{\bb{b}} = J^{-2/3} \bb{b} = J^{-2/3} \bb{F} \cdot \bb{F}^{\text{T}}
\end{eqnarray}
In the limit of small deformations $\lVert \nabla \mathbf{u} \rVert  \ll 1$, neo-Hookean hyperelasticity (Equation \ref{eq:neoHook}) reduces to linear elasticity (Equation \ref{eq:linearElastic}).

\subsection{Guccione Hyperelasticity}
The \citet{Guccione1995} hyperelastic law defines the second Piola-Kirchhoff stress as
\begin{eqnarray}
	\boldsymbol{S}
		%&=&  \frac{\partial \Psi}{\partial \boldsymbol{E} } \notag \\
		&=& \frac{\partial Q}{\partial \boldsymbol{E} } \left( \frac{C}{2} \right) e^Q + \frac{\kappa}{2} \frac{J^2 - 1}{J} \textbf{I} 
\end{eqnarray}
where
\begin{eqnarray}
%	\Psi(I_1, I_2, I_4, I_5) &=& \frac{C}{2} \left( e^{Q(I_1, I_2, I_4, I_5)} - 1 \right) \\
	Q(I_1, I_2, I_4, I_5) &=& c_t I_1^2 - 2 c_t I_2 + (c_f - 2c_{fs} + c_t) I_4^2   + 2(c_{fs} - c_t) I_5 \\
	\frac{\partial Q}{\partial \boldsymbol{E} }
%		&=& \frac{\partial Q}{\partial I_1} \frac{\partial I_1}{\partial \boldsymbol{E} } + \frac{\partial Q}{\partial I_2} \frac{\partial I_2}{\partial \boldsymbol{E} }
%		+ \frac{\partial Q}{\partial I_4} \frac{\partial I_4}{\partial \boldsymbol{E} } + \frac{\partial Q}{\partial I_5} \frac{\partial I_5}{\partial \boldsymbol{E} } \notag \\
		&=&  2 c_t \boldsymbol{E} + 2(c_f - 2 c_{fs} + c_t)I_4 ( \boldsymbol{f_0} \otimes \boldsymbol{f_0} ) \notag \\
		&&+ 2(c_{fs} - c_t)\left[\boldsymbol{E} \cdot  (\boldsymbol{f_0} \otimes \boldsymbol{f_0}) + ( \boldsymbol{f_0} \otimes \boldsymbol{f_0} ) \cdot \boldsymbol{E} \right]
\end{eqnarray}
The scalars $C$, $c_f$, $c_{fs}$, and $c_t$ are material parameters and invariants of the Green strain, $\boldsymbol{E} = \boldsymbol{F}^{\text{T}} \cdot \boldsymbol{F}$, are defined as
\begin{eqnarray}
	I_1 = \text{tr}(\boldsymbol{E}), \quad
	I_2 =  \frac{1}{2} \left[ \text{tr}^2(\boldsymbol{E}) - \text{tr}(\boldsymbol{E} \cdot \boldsymbol{E}) \right], \notag \\
	I_4 = \boldsymbol{E}  : \left( \boldsymbol{f_0} \otimes \boldsymbol{f_0} \right), \quad 
	I_5 = \left( \boldsymbol{E} \cdot \boldsymbol{E} \right): \left( \boldsymbol{f_0} \otimes \boldsymbol{f_0} \right)
\end{eqnarray}
with $\boldsymbol{f_0}$ representing the unit fibre directions in the initial configuration.

Equation \ref{eq:S2sigma} is used to convert the second Piola-Kirchhoff stress to the true stress.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Meshes Used with the Method of Manufactured Solutions}
\label{app:meshes}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[htbp]
	\centering
	\subfigure[Regular tetrahedral mesh with $4\,374$ cells]
	{
   		\includegraphics[width=0.45\textwidth]{figures/mms_mesh_tet}  
   	}
	\subfigure[Regular polyhedral mesh with $1\,000$ cells]
	{
   		\includegraphics[width=0.45\textwidth]{figures/mms_mesh_poly}  
   	}
	\subfigure[Regular hexahedral mesh with $1\,000$ cells]
	{
   		\includegraphics[width=0.45\textwidth]{figures/mms_mesh_hex}  
   	}
	\subfigure[Distorted hexahedral mesh with $1\,000$ cells]
	{
   		\includegraphics[width=0.45\textwidth]{figures/mms_mesh_hex_distorted}  
   	}
	\caption{Meshes used in the method of manufactured solutions case}
	\label{fig:mms_all}
\end{figure}
